---
title: "Income Census- Exploratory data analysis and classification"
author: "Minh Trung DANG"
date: "28/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Predicting if income exceeds $50,000 per year based on 1994 US Census Data with Simple Classification Techniques

Data Set Information:

Extraction was done by Barry Becker from the 1994 Census database. A set of reasonably clean records was extracted using the following conditions: ((AAGE>16) && (AGI>100) && (AFNLWGT>1)&& (HRSWK>0))
download the dataset from: "https://www.kaggle.com/uciml/incomecensus-census-income"


*** Response variable
A. income: Categorical variable that contains yearly income of the respondent ("<=$50K" or ">50K").

*** Independent variables
B. age: Numerical variable that contains the age of the respondent.

C. workclass : Categorical variable that contains the type of employer of the respondent

D. fnlwgt: Numerical variable that contains the number of respondents that each row of the data set represents.

E. education: Categorical variable that represents the level of education of the respondent (Doctorate, Prof-school, Masters, Bachelors, Assoc-acdm, Assoc-voc, Some-college, HS-grad, 12th, 11th, 10th, 9th, 7th-8th, 5th-6th, 1st-4th, Preschool)

F. education.num: Numerical variable that represents the *education* variable.

G. marital.status: The marital status of the respondent

H. occupation: Categorical variable that represents the type of employment of the respondent (?, Adm-clerical, Armed-Forces, Craft-repair, Exec-managerial, Farming-fishing, Handlers-cleaners, Machine-op-inspct, Other-service, Priv-house-serv, Prof-specialty, Protective-serv,  Sales,  Tech-support, Transport-moving). 

I. relationship: Categorical variable that represents the position in the family of the respondent (Husband, Not-in-family, Other-relative, Own-child, Unmarried, Wife).

J. race: Categorical variable that represents the race of the respondent (Amer-Indian-Eskimo, Asian-Pac-Islander, Black,Other, White).

K. sex: Categorical variable that represent the sex of the respondent (Female, Male).

L. capital.gain: Numerical variable that represents the income gained by the respondent from sources other than salary/wages.

M. capital.loss: Numerical variable that represents the income lost by the respondent from sources other than salary/wages.

N. hours.per.week: Numerical variable that represents the hours worked per week by the respondent.

O. native.country: Categorical variable that represents the native country of the respondent. 


1. Importing data

We will read in our data using the read_csv() function, from the tidyverse package readr, instead of read.csv().

```{r}
library(readr)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(DataExplorer)
library(pander)
```

```{r}
incomecensus <- read_csv("adultincome.csv")
```

2. Inspecting the data frame

First look at the data

```{r}
str(incomecensus)
```

```{r}
incomecensus <- incomecensus %>% 
  select(-c(fnlwgt))
```

```{r}
head(incomecensus)
```

Dealing with rows containing  "?"

We can see that there are some rows with "?".

For example

```{r}
incomecensus %>% 
  group_by(workclass) %>% 
  summarise(count = n())
```

We will check if there are any other columns containings "?"

First, check for the records with any "?"

```{r}
missing_count <- purrr::map_df(incomecensus, ~ stringr::str_detect(., pattern = "\\?")) %>%
  rowSums() %>%
  tbl_df() %>%
  filter(value > 0) %>%
  summarize(missing_count = n()) 

missing_count
```

In total, there are 2399 rows containing "?"

```{r}
count.NA.per.col <- plyr::ldply(incomecensus, function(c) sum(c == "?"))
count.NA.per.col %>% 
  pander()
```  

There are 3 columns that contain "?" as NA: workclass, occupation, and native.country

We will remove all of these records.

```{r}
incomecensus <- incomecensus %>% 
  filter(!workclass=="?", !occupation=="?", !native.country=="?")
```

The resulting data contains 30162 rows.

Assigning correct R data types to each column

Converting character to factor.m

```{r}
incomecensus <- incomecensus %>%
  mutate_if(is.character,as.factor)

str(incomecensus)
```

3. Detecting the inconsistency of the data

3.1. Missing Values

```{r}
plot_missing(incomecensus)
```

We will remove the fnlwgt column

3.2. Checking duplicate records

```{r duplicates}
    # to check for duplicate records
incomecensus %>%
  summarize(record_count = n(),
            distinct_records = n_distinct(.))
```

The function distinct() [dplyr package] can be used to keep only unique/distinct rows from a data frame. 

If there are duplicate rows, only the first row is preserved. It’s an efficient version of the R base function unique().

```{r}
incomecensus <- incomecensus %>% 
  distinct()
```

The resulting data frame contain 26904 records.


3.3. Checking the consistence of the data

A.  Age and education


```{r}
incomecensus %>%
  ggplot(aes(age, education)) + 
  geom_point(alpha = 0.3, col = "#00AFBB") + 
  geom_point(aes(col = (age<= 20 & education == 'Masters')), 
             alpha = 1, 
             size = 3) + 
  scale_colour_manual(values = setNames(c('blue','grey'),
                                        c(T, F))) +
  theme(legend.position="bottom") + 
  scale_x_continuous(breaks = seq(0,100,10))+ 
  xlab("Age") + 
  ylab("Education Level") +
  ggtitle("Age vs Education Level")
```
There are a few data points (in blue) are outlier datapoints.
 
It is highly unlikely that the people with age of 20 years and less can complete Masters degree. In reality, it might be possible.

We can take these data points out. 

```{r}
incomecensus <- incomecensus %>%
  filter(!(age <= 20 & education == 'Masters'))
```

B. sex and relationship status

```{r}
table(incomecensus$sex, incomecensus$relationship)
```

There is one record where Husband is female

There is also one record where wife is female. 

In 1994, this situation was not possible in the US

Where are these records ?

```{r}
incomecensus %>%
  filter((sex == 'Male' & relationship == 'Wife')  | (sex == 'Female' & relationship == 'Husband'))
```

These points can be removed using the code below.

```{r}
incomecensus <- incomecensus %>%
  filter(!((sex == 'Male' & relationship == 'Wife') | (sex == 'Female' & relationship == 'Husband')))
```

C. Relationship and marital status

Marital Status: The marital status classification identifies five major categories: never married, married, widowed, and divorced.

```{r}
table(incomecensus$relationship, incomecensus$marital.status) 
```

The group “other married, spouse absent” includes married people living apart because either the husband or wife was employed and
living at a considerable distance from home, was serving away from home in the Armed Forces, had moved to another area, or had a different place of residence for any other reason except separation as defined above.
(https://www.unmarried.org/government-terminology/)


```{r}
incomecensus %>%
  filter(marital.status == 'Married-spouse-absent' & relationship == 'Unmarried') %>% 
  summarise(count = n())
```

There are 120 rows where people are 'Married-spouse-absent' and 'Unmarried' at the same time.

Thers records will be removed.

```{r}
incomecensus <- incomecensus %>%
  filter(!(marital.status == 'Married-spouse-absent' & relationship == 'Unmarried'))
```


D. age and weekly working hour

```{r, fig.width= 6, fig.height= 6}
incomecensus %>%
  ggplot(aes(age, hours.per.week)) + 
  geom_point(size = 2, shape = 23, color = "blue") + 
  theme(legend.position="bottom") + 
  scale_x_continuous(breaks = seq(0,100,10))+ 
  scale_y_continuous(breaks = seq(0,100,10))+ 
  xlab("Age") + 
  ylab("Weekly working hours") +
  ggtitle("Age vs Weekly working hours ")
```
Some people aged older than 70 and work more than 80 hours a week.

Even for people with 90 years of age, there are records of 100 hours per week.

To lighlight these data point


```{r, fig.width= 6, fig.height= 6}
incomecensus %>%
  ggplot(aes(age, hours.per.week)) + 
  geom_point(alpha = 0.3, col = "#00AFBB") +
  geom_point(aes(col = (age >= 70 & hours.per.week >= 40)), 
             alpha = 1, 
             size = 3) + 
  scale_colour_manual(values = setNames(c('red','grey'),
                                        c(T, F))) +
  theme(legend.position="bottom") + 
  scale_x_continuous(breaks = seq(0,100,10))+ 
  scale_y_continuous(breaks = seq(0,100,10))+ 
  xlab("Age") + 
  ylab("Weekly working hours") +
  ggtitle("Age vs Weekly working hours ")
```

```{r}
incomecensus %>%
  filter(age >= 70 & hours.per.week > 50) %>% 
  summarise(counts =n())
```

Data points with age more than and equal to 70 years and working hours greater than 50 hours can be removed.

```{r}
incomecensus <- incomecensus %>%
   filter(!(age >= 70 & hours.per.week > 50))
```


E. capital loss and capital gain

There are many points with 99999 as the Capital which seems suspicious and hence, should be eliminated from the analysis.

```{r}
incomecensus %>%
  group_by(capital.gain) %>% 
  summarise(counts =n()) %>% 
  arrange(desc(capital.gain)) %>% 
  head(10)
```

We can remove all the records where the capital gain is equal to 99999.

```{r}
incomecensus <- incomecensus %>%
   filter(!(capital.gain == 99999))
```


```{r}
incomecensus %>%
  group_by(capital.loss) %>% 
  summarise(counts =n()) %>% 
  arrange(desc(capital.loss)) %>% 
  head(10)
```

F. Work class and occupation

```{r}
table(incomecensus$occupation, incomecensus$workclass)
```

We might remove the two levels: without-pay 

We also might regroup these levels into 3 categoris : public, private, self-employeed 


3. Exploratory data analysis (EDA)

***A. income***

Categorical variable that contains yearly income of the respondent ("<=$50K" or ">50K").

```{r}
incomecensus %>% 
  group_by(income) %>% 
  summarise(count = n())
```

```{r}
income_prop <- incomecensus %>% 
  group_by(income) %>% 
  summarise(count = n()) %>% 
  ungroup()%>% 
  arrange(desc(income)) %>%
  mutate(percentage = round(count/sum(count),4)*100,
         lab.pos = cumsum(percentage)-0.5*percentage)

ggplot(data = income_prop, 
       aes(x = "", 
           y = percentage, 
           fill = income))+
  geom_bar(stat = "identity")+
  coord_polar("y") +
  geom_text(aes(y = lab.pos, 
                label = paste(percentage,"%", sep = "")), col = "blue", size = 5) +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  theme_void() +
  theme(legend.title = element_text(color = "black", size = 14),
        legend.text = element_text(color = "black", size = 14))

```

The summary of the data shows that  75 % of the observations have an income less than or equal to 50k dollars. 

Specifically, 22654 persons have an income <=50k dollars, while  7508 people earn more than 50k. 

***B. Age***
```{r}
incomecensus %>% ggplot(aes(age)) + 
  geom_histogram(fill= "lightblue",
                 color = 'blue',
                 binwidth = 5) +   
  labs(title= "Age Distribution") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
summary(incomecensus$age)
```

```{r}
incomecensus %>% ggplot(aes(age)) + 
  geom_histogram(aes(fill=income),
                 color = 'grey',
                 binwidth = 1) +   
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  labs(title= "Age Distribution for Income")+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r age, echo=TRUE, eval=TRUE}
incomecensus %>% 
  ggplot(aes(age, 
             fill= income)) +
  geom_density(alpha= 0.7, color = 'blue') +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  labs(x = "Age", y = "Density", title = "Density graph of age distribution")
```

Older people  tends to earn more.

On the other hand, majority of the people with an age around 25 years earns less than 50k per annum.

***C. workclass ***
```{r}
library(ggthemes)
library(hrbrthemes)
```

```{r}
incomecensus %>%
  group_by(workclass) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(desc(counts)) %>% 
  ggplot(aes(x= reorder(workclass, -counts),
             y  = counts,
             fill = Percentage)) +
  geom_bar(stat = "identity",
           width = 0.6) +
  scale_fill_gradient(low="skyblue1", high="royalblue4")+
  geom_text(aes(label = paste0(round(counts,1), "\n",Percentage,"%")), 
            vjust = -0.1, 
            color = "darkblue", 
            size = 4) +
  scale_y_continuous(limits = c(0,25000)) +
  theme_minimal() +
  labs(x = "Work class",
       y = "Frequency",
       caption = "Income census US 1994") 
```

Almost three-quarters of sample work in the private sector. 

In this data, there are dictinction between Local-gov, State-gov, and Federal-gov.

We might group these three category if there is no difference in income level between them.

```{r}
library(ggpubr)
library(scales)
```

```{r, fig.height= 7.5, fig.width= 20}
p1 <- incomecensus %>%
  group_by(workclass) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(desc(counts)) %>% 
  ggplot(aes(x= reorder(workclass, counts),
             y  = counts)) +
  geom_bar(stat = "identity",
           width = 0.6,
           fill = "steelblue") +
  geom_text(aes(label = paste0(round(counts,1),"\n",Percentage,"%")), 
            vjust = 0.5, 
            hjust = -0.5,
            color = "darkblue", 
            size = 4) +
  scale_y_continuous(limits = c(0,20000)) +
  theme_minimal() +
  labs(x = "Work class",y = "Frequency") + 
  coord_flip()

p2 <- incomecensus %>% 
  group_by(workclass, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(workclass, n), 
             y = pct/100, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_y_continuous(name= "Percentage", 
                     labels = percent) +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.y=element_blank()) + 
  coord_flip()

ggarrange(p1, p2, nrow = 1)
```

````{r}
income_gov <- incomecensus %>% 
  filter(workclass %in% c("Local-gov", "State-gov", "Federal-gov")) %>% 
  group_by(workclass, income) %>%
  summarise(count = n()) %>% 
  mutate(pct = count/sum(count)) %>%
  arrange(desc(income), pct)

income_gov <-  income_gov %>% transmute(income, percent = count*100/sum(count))
income_gov 
```  

```{r, fig.height= 5, fig.width= 5}
incomecensus %>% 
  filter(workclass %in% c("Local-gov", "State-gov", "Federal-gov")) %>% 
  group_by(workclass, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(workclass, n), 
             y = pct, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

There is a slight difference between Local-gov and Stave-gov. 

Working for the federal government can have a greater chance of getting an income higher than 50K.

```{r, fig.height= 5, fig.width= 5}
incomecensus %>% 
  filter(workclass %in% c("Self-emp-not-inc", "Self-emp-inc", "Without-pay")) %>% 
  group_by(workclass, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(workclass, n), 
             y = pct, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```

***education***
```{r}
incomecensus %>% 
  group_by(education) %>% 
  summarise(counts = n()) %>% 
  arrange(desc(counts)) %>% 
  pander()
```

```{r}
incomecensus %>% 
  group_by(education.num) %>% 
  summarise(counts = n()) %>% 
  arrange(desc(counts)) %>% 
  pander()
```
We can see that these two columns are identical. 

The column education.num is simply converted from the column education.

If we keep the column education which is factor type, we might regroup some categories in this column. 

If we keep the column education.num which is numerical type,, we can keep the column as it is (numeric).

```{r}
eudcation_df <- incomecensus %>%
  group_by(education.num) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(education.num)

eudcation_df %>% 
  pander()
```

```{r,, fig.height= 5, fig.width= 12}
ggplot(data = eudcation_df ,
           aes(x= education.num,
               y  = counts,
               fill = Percentage)) +
  geom_bar(stat = "identity",
           width = 0.6) +
  scale_x_continuous(breaks = c(0:16)) +
  scale_fill_gradient(low="skyblue1", high="royalblue4")+
  geom_text(aes(label = paste0(round(counts,1), "\n","(",Percentage,"%)")), 
            vjust = -0.1, 
            color = "darkblue", 
            size = 4) +
  scale_y_continuous(limits = c(0,12000)) +
  theme_minimal() +
  labs(x = "Years of education",
       y = "Frequency",
       caption = "Income census US 1994") +
  ggtitle("Distribution of years of education") +
    theme(plot.title = element_text(hjust = 0.5), 
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(colour="black", size = 12))
```

About 50% of respondents have 9-10 years of education.

About 17% of people in the sample hold bachelor degree (13 years of education). 


```{r, fig.height= 5, fig.width= 7.5}
education.pct <- incomecensus %>%
  group_by(education.num, income) %>%
  summarize(count = n()) %>%
  mutate(pct = count/sum(count)) %>%
  arrange(desc(income), pct)

ggplot(education.pct, 
       aes(education.num, pct, fill = income)) + 
  geom_bar(stat="identity", position = "fill") + 
  geom_hline(yintercept = 0.2489, col = "blue") +
  scale_x_continuous(breaks = c(0:16)) +
  scale_y_continuous(labels=scales::percent) + 
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) + 
  ggtitle("Income by years of education") + 
  xlab("Education (years)") + 
  ylab("Percentage") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(colour="black", size = 12)) 
```
Higher years of education led to a higher chance of having >50K.

***marital status***
```{r}
incomecensus %>% 
  group_by(marital.status) %>% 
  summarise(counts = n()) %>% 
  arrange(desc(counts))
```
we can put all the married people in the same group

This column have 7 labels, three of them are:

- Married-AF-spouse: Married armed forces spouse  

- Married-civ-spouse: Married civilian spouse  

- Married-spouse-absent

These levels can be grouped into the group "married".  

Replace  "Married-AF-spouse", "Married-civ-spouse", and "Married-spouse-absent" by "Married"

```{r}
pat = c("Married-AF-spouse|Married-civ-spouse|Married-spouse-absent")
incomecensus <- incomecensus %>% 
  mutate(marital_status = stringr::str_replace_all(marital.status, pat, "Married"))
```

```{r}
incomecensus <- incomecensus %>% 
  select(-marital.status)
```

```{r}
marital_df <- incomecensus %>%
  group_by(marital_status) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),1)) %>% 
  arrange(desc(counts))

marital_df %>% 
  pander()
```

The proportion of married people that have an income higher than 50K is highest. 

```{r, fig.height= 4, fig.width= 12}
p3 <- incomecensus %>%
  group_by(marital_status) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(desc(counts)) %>% 
  ggplot(aes(x= reorder(marital_status, counts),
             y  = counts)) +
  geom_bar(stat = "identity",
           width = 0.6,
           fill = "steelblue") +
  geom_text(aes(label = paste0(round(counts,1),"\n","(",Percentage,"%)")), 
            vjust = 0.4, 
            hjust = -0.5,
            color = "darkblue", 
            size = 4) +
  scale_y_continuous(limits = c(0,20000)) +
  theme_minimal() +
  labs(x = "Marital status",y = "Frequency") + 
  coord_flip()

p4 <- incomecensus %>% 
  group_by(marital_status, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(marital_status, n), 
             y = pct/100, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_y_continuous(name = "Percentage", 
                     labels = scales::percent) +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.y=element_blank()) + 
  coord_flip()

ggarrange(p3, p4, nrow = 1)
```


***occupation***

```{r, fig.height= 7.5, fig.width= 10}
Category_df <- incomecensus %>%
  group_by(occupation) %>% 
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(counts)

ggplot(data = Category_df,
           aes(x= reorder(occupation, -counts),
               y  = counts,
               fill = Percentage)) +
  geom_bar(stat = "identity",
           width = 0.7) +
  geom_text(aes(label = paste0(counts,", ", Percentage,"%")), 
            vjust = 0.2, 
            hjust = -0.1,
            color = "darkblue", 
            size = 4) +
  scale_y_continuous(limits = c(0,5000)) +
  scale_fill_distiller(palette = "Spectral") +
  theme_minimal() +
  labs(x = "Category",
       y = "Frequency",
       caption = "Income census US 1994") +
  coord_flip()
```

```{r, fig.height= 6, fig.width= 7.5}
occupation.pct <- incomecensus %>%
  group_by(occupation, income) %>%
  summarize(count = n()) %>%
  mutate(pct = count/sum(count)) %>%
  arrange(desc(income), pct)

occupation.pct$occupation <- factor(occupation.pct$occupation,
                                    levels = occupation.pct$occupation[1:(nrow(occupation.pct)/2)])

ggplot(occupation.pct, aes(reorder(occupation,-pct), pct, fill = income)) + 
  geom_bar(stat="identity", position = "fill") + 
  geom_hline(yintercept = 0.2489, col = "blue") +
  ggtitle("Income by occupation") + 
  xlab("Occupation") + 
  ylab("Percentage") + 
  scale_y_continuous(labels=scales::percent) + 
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12),
        legend.position = "bottom", 
        legend.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(colour="black", size = 12)) + 
  coord_flip()
```

About 25% of the sample has an income of more than 50K. 

There are five categories of jobs that have a higher percentage: Protective service, tech-support, Sales, Exec-managerial, and Prof-speciality.

Question: Are these people work in the private sector?

```{r}
incomecensus %>% 
  filter(occupation %in% c("Protective-serv", "Tech-support", "Sales", "Exec-managerial", "Prof-specialty")) %>% 
  group_by(workclass) %>% 
  summarise(freq =  n()) %>% 
  mutate(percentage = round(freq*100/sum(freq),1)) %>%
  arrange(desc(percentage)) %>% 
  pander()
```

About two-third work in the private sector.

Governmental job accounts for 18.81 %.

Blue_Collar (Craft-repair,Farming-fishing,Handlers-cleaners,Machine-op-inspct,Transport-moving)

White_Collar (Adm-clerical, Sales, Tech-support,Protective-serv), 

Exec_mgr_prof (Exec-managerial,Prof-specialty), 

and Service_other (Armed-Forces, Other-service, Priv-house-serv)

```{r ,echo=FALSE}
incomecensus$occupation <- incomecensus$occupation %>% 
  fct_collapse(Blue_collar=c("Craft-repair","Farming-fishing","Handlers-cleaners","Machine-op-inspct","Transport-moving"),
               White_collar=c("Adm-clerical","Sales","Tech-support","Protective-serv"),
               Exec_mgr_prof=c("Exec-managerial","Prof-specialty"), 
               Other_service =c("Armed-Forces", "Other-service", "Priv-house-serv"))
```

```{r,fig.height= 4, fig.width= 12}
occupation_distr <- incomecensus %>%
  group_by(occupation) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(desc(counts)) %>% 
  ggplot(aes(x= reorder(occupation, counts),
             y  = counts)) +
  geom_bar(stat = "identity",
           width = 0.6,
           fill = "steelblue") +
  geom_text(aes(label = paste0(round(counts,1),"\n","(",Percentage,"%)")), 
            vjust = 0.4, hjust = -0.5,color = "darkblue", size = 4) +
  scale_y_continuous(limits = c(0,20000)) +
  theme_minimal() +
  labs(x = "Grouped Occupation", y = "Frequency") + 
  coord_flip()

occupation_pct <- incomecensus %>% 
  group_by(occupation, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(occupation, n), 
             y = pct/100, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_y_continuous(name = "Percentage", 
                     labels = scales::percent) +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.y=element_blank()) + 
  coord_flip()

ggarrange(occupation_distr, occupation_pct, nrow = 1)
```

```{r relationship, echo=TRUE, eval=TRUE}
incomecensus %>% 
  ggplot(aes(income, color= income, fill= income)) +
  geom_bar( alpha = 0.8, width = 0.8) +
  facet_grid(~ relationship) + 
  labs(x = "Incomes", y = "Count", title = "Incomes by relationship")
```

This column contains categories that might be overlap with other feature. 

For example, unmarried people in the relationship attribute is the same as the unmarried level in the marital column.


```{r,fig.height= 4, fig.width= 12}
relationship_distr <- incomecensus %>%
  group_by(relationship) %>%
  summarise(counts = n()) %>% 
  mutate(Percentage = round(counts*100/sum(counts),2)) %>% 
  arrange(desc(counts)) %>% 
  ggplot(aes(x= reorder(relationship, counts),
             y  = counts)) +
  geom_bar(stat = "identity",
           width = 0.6,
           fill = "steelblue") +
  geom_text(aes(label = paste0(round(counts,1),"\n","(",Percentage,"%)")), 
            vjust = 0.4, hjust = -0.5,color = "darkblue", size = 4) +
  scale_y_continuous(limits = c(0,20000)) +
  theme_minimal() +
  labs(x = "Relationship", y = "Frequency") + 
  coord_flip()

relationship_pct <- incomecensus %>% 
  group_by(relationship, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(relationship, n), 
             y = pct/100, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_y_continuous(name = "Percentage", 
                     labels = scales::percent) +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.y=element_blank()) + 
  coord_flip()

ggarrange(relationship_distr, relationship_pct, nrow = 1)
```

This variable may be redundant as it collapse somehow with the feature marital_status.


***race***

```{r, fig.height = 4, fig.width= 7.5}
incomecensus %>% 
  ggplot(aes(race, fill= income)) +
  geom_bar(position = "fill") +
  labs(x = "Race", 
       y = "Proportion", 
       title = "Incomes by race")+
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_hline(yintercept = 0.2489, col="blue") +
  coord_flip()
```
Black and "Amer-Indio-Eskima" earn less than 50k more frequently than the general population. 

However, the percent of whites and "Asian-Pac-Islander" that earn more than 50k is over the general population average.

***Sex***
Here we can see that the vast majority of people having an income greater than 50000 dollars are males.

```{r}
gender_prop <- incomecensus %>% 
  group_by(sex) %>% 
  summarise(count = n()) %>% 
  ungroup()%>% 
  arrange(desc(sex)) %>%
  mutate(percentage = round(count/sum(count),4)*100,
         lab.pos = cumsum(percentage)-0.5*percentage)

gender_distr <- ggplot(data = gender_prop, 
       aes(x = "", 
           y = percentage, 
           fill = sex))+
  geom_bar(stat = "identity")+
  coord_polar("y") +
  geom_text(aes(y = lab.pos, 
                label = paste(percentage,"%", sep = "")), col = "blue", size = 4) +
  scale_fill_manual(values=c("orange", "lightblue"),
                    name = "Gender") +
  theme_void() +
  theme(legend.title = element_text(color = "black", size = 12),
        legend.text = element_text(color = "black", size = 12))


gender_prop <- incomecensus %>% 
  group_by(sex, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(sex, n), 
             y = pct, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "black") +
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(color = "black", size = 12),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.title = element_text(color = "black", size = 12),
        legend.text = element_text(color = "black", size = 12))

ggarrange(gender_distr, gender_prop, nrow = 1)
```
Two-third of respondents are male. 

The proportion of women that earn more than 50k is much lower than that of their male counterparts.

***hour per week***

Distribution of hours per week

```{r}
incomecensus %>% ggplot(aes(hours.per.week)) + 
  geom_histogram(fill= "orange",
                 color = 'blue',
                 binwidth = 5) +   
  labs(title= "Age Distribution") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r workw, echo=TRUE, eval=TRUE}
incomecensus %>% 
  group_by(income) %>%
  summarise("Mean hours per week" = mean(hours.per.week),
            "Standard deviatrion" = sd(hours.per.week))
```

Contrusting box-plot

```{r, fig.height= 5, fig.width = 5}
ggplot(data = incomecensus, 
       aes(income, 
           hours.per.week, 
           fill = income))+
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) + 
  geom_boxplot()+
  labs(x = "Incomes", 
       y = "Worked hours per week", 
       title = "Incomes by working hours")
```

There are many outliers. 

Some people work for 100 hours a week (which is possible)

***native.country***

Incomes by origin 

There are 40 different  native countries and some of them have just a few cases, for instance, honduras have just 10 cases.

Therefore, the first attempt is to group the countries by continent. 

```{r}
incomecensus <- incomecensus %>% 
  mutate(native_continent = case_when(
    native.country %in% c("France", "Greece", "Hungary", "Italy", "Portugal",
                          "Scotland", "England", "Germany", "Holand-Netherlands",
                          "Ireland", "Poland", "Yugoslavia") ~ "Europe",
    native.country %in% c("Columbia", "Dominican-Republic", "El-Salvador", "Haiti",
                          "Honduras", "Mexico", "Outlying-US(Guam-USVI-etc)",
                          "Cuba", "Ecuador", "Guatemala", "Jamaica", "Nicaragua",
                          "Peru", "Puerto-Rico", "Trinadad&Tobago") ~ "Latin America", # "Trinadad&Tobago" and "Outlying-US(Guam-USVI-etc" 
    native.country %in% c("Iran", "Japan", "Philippines", "Taiwan", "Vietnam", "Cambodia",
                          "China", "Hong", "India", "Laos", "South", "Thailand") ~ "Asia",
    native.country %in% c("United-States","Canada") ~ "USA/Canada"))
```


```{r}
incomecensus %>% 
  group_by(native_continent) %>% 
  count()
```

```{r, fig.width=6 , fig.height= 5}
incomecensus %>% 
  group_by(native_continent, income) %>% 
  summarize(n = n()) %>% 
  mutate(pct = n*100/sum(n)) %>% 
  ggplot(aes(x = reorder(native_continent, n), 
             y = pct, 
             fill = income)) +
  geom_bar(stat = "identity", width = 0.6) + 
  scale_x_discrete(name = "") +
  scale_fill_manual(values=c("#E3CD81FF", "#B1B3B3FF")) +
  geom_text(aes(label = paste0(round(pct,0),"%")), 
            position = position_stack(vjust = 0.5), 
            size = 4, 
            color = "blue") +
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(color = "black", size = 12),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.title = element_text(color = "black", size = 12),
        legend.text = element_text(color = "black", size = 12))
```


We will remove some columns

```{r}
incomecensus <- incomecensus %>% 
  select(-c("education","capital.gain","capital.loss", "relationship", "native.country"))
```



The final data 


```{r}
str(incomecensus)
```
we need to convert two columns marital_status and native_continent in factor

```{r}
incomecensus$marital_status <- as.factor(incomecensus$marital_status)
incomecensus$native_continent <- as.factor(incomecensus$native_continent)
```

we will save this data as income.csv



```{r}
write.csv(incomecensus,file = "income.csv", row.names= FALSE)
```



MODELLLING

```{r}
data <- read_csv("income.csv")
```

```{r}
data <- data %>%
  mutate_if(is.character,as.factor)

str(data)
```

```{r}
library(caret)
```


```{r}
Trainindex <- createDataPartition(y = data$income , p = .70, list = FALSE)

training <- data[Trainindex ,]

validation <- data[-Trainindex,]

training_new <- training[-8]

validation_new <- validation[-8]

income_training_label <- training$income

income_validation_label <- validation$income
```


***LOGISTIC REGRESSION***
package caret

```{r}
set.seed(123)
default_glm_mod <- train(form = income ~ .,
                         data = training,
                         method = "glm",
                         family = "binomial",
                         tuneLength = 5)
```

```{r}
glm_pred <- predict(default_glm_mod, newdata = validation)

confusionMatrix(glm_pred, validation$income)
```


***II. Decision tree based methods***

II.1 rpart

Fully grown trees

```{r}
library(rpart)
library(rpart.plot)
library(caret)
```


```{r}
rparttree <- rpart(income ~ ., 
                   data = training,
                   method = "class")
```

```{r}
rparttree 
```


```{r, fig.height = 7.5, fig.width = 10}
# Plot the trees
rpart.plot(rparttree)
```

Marital status is the most important feature for this data set.

```{r}
plotcp(rparttree)
```


II.1. C5.0 with boosting

```{r}
library(C50)
```

```{r}
set.seed(123)

treeC5 <- C5.0(income ~ .,
                data = training,
                trials = 100)

treeC5_Pred <- predict(treeC5, validation)

confusionMatrix(treeC5_Pred, validation$income)
```

II.2 Randomm forest

```{r}
library(randomForest)
```

```{r}
set.seed(123)
treeRf <- randomForest(income ~ .,
 data = training,
 ntree = 500,
 mtry = 3,
 importance = TRUE)
```

```{r}
treeRf_Pred <- predict(treeRf, validation)

confusionMatrix(treeRf_Pred, validation$income)
```

***NAIVE BAYES***


```{r}
library(e1071)
```

```{r}
set.seed(123)
NB <- naiveBayes(income ~., data = training)
# that one is faster than the package caret
```


```{r}
NB_pred <- predict(NB, validation, type="class")
confusionMatrix(NB_pred,validation$income)
```

***KNN***
```{r}
set.seed(3333)
knn_fit <- train(income ~., 
                 data = training, 
                 method = "knn",
                 preProcess = c("center", "scale"),tuneLength = 5)
```


```{r}
knn_fit
```

```{r}
KNN_pred <- predict(knn_fit, validation, type="raw")
confusionMatrix(KNN_pred,validation$income)
```





























